Combining Object-Oriented Paradigm and Controlled Natural Language for Requirements Specification

### 2021-09-13 article.tex
clean
polished

\title{Combining Object-Oriented Paradigm and Controlled Natural Language for Requirements Specification}
\author{Yegor Bugayenko\\yegor256@gmail.com}
\begin{abstract} Natural language is the dominant form of writing software requirements.
Its essential ambiguity causes inconsistency of requirements,
which leads to scope creep. On the other hand, formal requirements
specification notations such as Z, \nospell{Petri Nets}, SysML, and others are
difficult to understand by non-technical project stakeholders. They often
become a barrier between developers and requirements providers. The article
presents a controlled natural language that looks like English but
is a strongly typed object-oriented language compiled to
UML/XMI. Thus, it is easily understood, at the same time, by non-technical
people, programmers, and computers. Moreover, it is formally verifiable and
testable. It was designed, developed, and tested in three commercial
software projects in order to validate the assumption that object-oriented
design can be applied to requirements engineering at the level of
specifications writing. The article outlines key features of the language and summarizes
the experience obtained during its practical application. \end{abstract}
\maketitle

@@

Software Requirements Specification (SRS) should be correct, unambiguous,
complete, consistent, ranked for importance and/or stability, verifiable,
modifiable, and traceable~\citep{ieee830}. Many times, it was proven that an SRS
document that lacks some of these properties has a high likelihood of
becoming a root cause of scope and schedule problems in a
project~\citep{wiegers03,ears09,chen09}.

Natural Language (NL) is the main presentation means in industrial
requirements documents~\citep{kof10,sinha10}, mostly due to its flexibility.
Very often, contributors to requirements specifications are not experts in
requirements engineering (RE). Impreciseness and flexibility of NL, being
its advantage in informal human communications, turns into a serious drawback
in RE. Problems that appear in NL requirement documents include
ambiguity, vagueness, complexity, duplication, wordiness, and
\nospell{untestability}~\citep{ears09}.

@@

Second, they all require sequential flow of language constructs, while in
modern agile software projects wiki pages is the most popular storage for
requirements documentation, instead of files or databases. Wiki pages, by
definition, are not sorted and may contain pieces of requirements
documentation that should be combined into a complete document in a random
order.

@@

To overcome these drawbacks, a new language was introduced,
implemented, and tested. The Sec.~\ref{sec:Requs} of
the article presents the language and gives a simple example of its use,
specifying requirements of a sample software system.
The Sec.~\ref{sec:lexical} briefly outlines lexical rules of the language.
The Sec.~\ref{sec:syntax} describes the syntax using extended \nospell{Backus-Naur}
form and gives a few practical examples. The Sec.~\ref{sec:semantic} outlines
a semantic layer of the compiler that is intended to detect vast majority
of inconsistencies in a requirements document.
The Sec.~\ref{sec:uml} gives an example of to-UML mapping implemented
by the compiler.
The Sec.~\ref{sec:ambiguity}
presents a numeric metric that measures ambiguity in a requirements
specification and could be used as a guidance for system analysts.
The Sec.~\ref{sec:results} presents empirical results obtained from three
commercial Java web projects over the last few years.
The Sec.~\ref{sec:conclusion} presents conclusions and observations made by
the author.

@@

It is an object-oriented CNL that resembles English and, because of that,
can be read and understood by compilers, system analysts, end-users, and
business people. The Fig.~\ref{fig:example} presents an example of a short
requirements definition document that describes a calculator that asks its
user for two floating point numbers and returns the result of their
division.

@@

The Fig.~\ref{fig:java} contains a close equivalent program in Java.

@@

\begin{grammar}
<class> ::= `[A-Z][A-Za-z]+'

<word> ::= `[a-z]+'

<use-case-ID> ::= `UC[0-9\\.]+'

<flow-ID> ::= `[0-9]+'

<informal> ::= `"([^"]|\\")+"'
\end{grammar}

In the Fig.~\ref{fig:example}, the entities \ff{SuD}, \ff{Fraction}, \ff{Float}, and
\ff{User} are classes. They start with capital letters and contain only
letters (both lowercase and capital). The \ff{user}, the \ff{fraction}, and
the \ff{quotient} are the variables bound to instances of classes.

Besides, there are a few reserved words, which may have special meaning in
certain places of the text, including \ff{the}, \ff{a}, \ff{includes},
\ff{requires}, etc.
A full list of them is defined below in the Sec.~\ref{sec:syntax}.

@@

\begin{grammar}
<SRS> ::= <clause>+

<clause> ::= <class-declaration> \\
| <class-construction> \\
| <method-declaration> \\
| <alternative-flow-declaration>
\end{grammar}

@@

\begin{grammar}
<class-declaration> ::= <class> <is-a> <parent-class>

<parent-class> ::= <informal> | <class>

<is-a> ::= `is' ( `a' | `an' )
\end{grammar}

The second line in the Fig.~\ref{fig:example} declares the class \ff{Fraction}
without any super-class but with an informal description
\ff{"math calculator"}.

@@

\begin{grammar}
<class-construction> ::= <class> <includes> `:' <slots>

<includes> ::= `includes' | `needs'

<slots> ::= <slot> ( `,' <slot> )*

<slot> ::= <variable> ( `as' <class> )?

<variable> ::= <word> `-s'?
\end{grammar}

The Fig.~\ref{fig:example} defines constructor arguments only for
the \ff{Fraction} class. The class \ff{User} don't have any arguments. The class
\ff{SuD} is a pre-defined class without arguments. The class \ff{Float} is a
pre-defined class with one argument.

@@

\begin{grammar}
<method-declaration> ::= <use-case-ID> `where' \\
<class> <signature> `:' <flows>

<signature> ::= ( <word>+ | <informal> ) \\
<subject>? <using>?

<subject> ::= <class> <binding>? | `the' <variable>

<binding> ::= `(' `a' <variable> `)'

<using> ::= `using' <subject> ( `and' <subject> )*

<flows> ::= <flow> ( `;' <flow> )+

<flow> ::= <flow-ID> `.' \\
( `The' <variable> <signature>
| `Fail' `since' <informal> )
\end{grammar}

@@

\begin{grammar}
<alternative-flow-declaration> ::= <use-case-ID> `/' \\
<flow-ID> `when' <informal> `:' <flows>
\end{grammar}

@@

Every method call in every use case should either
\begin{inparaenum}[\itshape a\upshape)]
\item be informal (the 8th, 9th, and 11th lines in the example at the Fig.~\ref{fig:example}),
\item formally refer to the method declared in a class, or
\item construct an instance of a class using a built-in method \ff{creates} (the 7th line in the example).
\end{inparaenum}

@@

Every method in the object-oriented way may throw an exception, using a
build-in method \ff{Fail since}. The method has one parameter of the class
\ff{Text}, which explains the reason. In the example in
the Fig.~\ref{fig:example}, method ``calculates'' of the class \ff{Fraction} may
throw an exception and it is handled by an alternative flow. In the language, all
exceptions are checked, which means that they should be handled explicitly,
by means of alternative flows of use cases.

@@

When semantic analysis is done the specification looks like an object of
the class \ff{SuD} that encapsulates other objects and methods. This object
hierarchy is convertible to UML diagrams~\citep{uml}. Specification in
the Fig.~\ref{fig:example} would be converted to the following UML diagrams
(to save space, the list is limited to the three most interesting
diagrams):

\begin{itemize}
\item Class diagram, in the Fig.~\ref{fig:uml-class}
\item Object diagram, in the Fig.~\ref{fig:uml-object}
\item Sequence diagram of ``divides two numbers'' method, in the Fig.~\ref{fig:uml-seq}
\end{itemize}

@@

Every UML diagram is a formal interoperable document in XMI~\citep{xmi} that
can be rendered in web, in \LaTeX, in PDF, or translated to Java or any
other object-oriented language. The Fig.~\ref{fig:xmi} shows an example XMI
for the class diagram.

@@

In the Fig.~\ref{fig:example}, ambiguity equals to $3/4$ since there are four
methods in total and three of them are defined informally.

A method is formally defined when its signature consists of one specially
reserved verb \ff{creates} and an optional list of constructor arguments,
as in the seventh line of the Fig.~\ref{fig:example}. A formally defined
method doesn't have any ambiguity since it is absolutely clear what is
expected as its input and output. All it does is instantiate a new object
and bind it to a variable. In Java that would mean calling a constructor.

The higher the ambiguity, the more system analysis is required for the
requirements document to make it unambiguously understandable by everybody
in a project. The ultimate goal of a system analyst is to break down
requirements until a target ambiguity is reached. In the commercial projects
explained in the Sec.~\ref{sec:results}, the $A$ metric was used for the planning
of RE activities. Every iteration a group of RE specialists had a goal of
decreasing the ambiguity until it reached a given value. It was not clear at
the beginning how much work that would take, but soon the RE team
obtained that knowledge and became capable of estimating
complexity of work using $A$ metric.

>>>>

\title{Combining Object-Oriented Paradigm and Controlled Natural Language for Requirements Specification}
\author{Yegor Bugayenko}{}{}
\email{yegor256@gmail.com}
\affiliation{\institution{}\city{Moscow}\country{Russia}}
\ccsdesc[100]{Requirements Engineering}
\keywords{Requirements, Natural Language Processing}

\begin{abstract} Natural language is the dominant form of writing software requirements.
Its essential ambiguity causes inconsistency of requirements,
which leads to scope creep. On the other hand, formal requirements
specification notations such as Z, \nospell{Petri Nets}, SysML, and others are
difficult to understand by non-technical project stakeholders. They often
become a barrier between developers and requirements providers. The article
presents a controlled natural language that looks like English but
is a strongly typed object-oriented language compiled to
UML/XMI. Thus, it is easily understood, at the same time, by non-technical
people, programmers, and computers. Moreover, it is formally verifiable and
testable. It was designed, developed, and tested in three commercial
software projects in order to validate the assumption that object-oriented
design can be applied to requirements engineering at the level of
specifications writing. The article outlines key features of the language and summarizes
the experience obtained during its practical application. \end{abstract}
\maketitle

@@

Software Requirements Specification (SRS) should be correct, unambiguous,
complete, consistent, ranked for importance and/or stability, verifiable,
modifiable, and traceable~\citep{ieee830}. Many times, it was demonstrated that an SRS
document that lacks some of these properties has a high likelihood of
becoming a root cause of scope and schedule problems in a
project~\citep{wiegers03,ears09,chen09}.

Natural Language (NL) is the main presentation mean in industrial
requirements documents~\citep{kof10,sinha10}, mostly due to its flexibility.
Very often, contributors to requirements specifications are not experts in
requirements engineering (RE). Impreciseness and flexibility of NL, being
its advantage in informal human communications, turns into a serious drawback
in RE. Problems that appear in NL requirement documents include
ambiguity, vagueness, complexity, duplication, wordiness, and
\nospell{untestability}~\citep{ears09}.

@@

Second, they all require sequential flow of language constructs, while in
modern Agile software projects wiki pages is one of the most popular storage for
requirements documentation, instead of files or databases. Wiki pages, by
definition, are not sorted and may contain pieces of requirements
documentation that should be combined into a complete document in a random
order.

@@

To overcome these drawbacks, a new language was introduced,
implemented, and tested.
Sec.~\ref{sec:Requs} of
the article presents the language and gives a simple example of its use,
specifying requirements of a sample software system.
Sec.~\ref{sec:lexical} briefly outlines lexical rules of the language.
Sec.~\ref{sec:syntax} describes the syntax using extended \nospell{Backus-Naur}
form and gives a few practical examples.
Sec.~\ref{sec:semantic} outlines
a semantic layer of the compiler that is intended to detect vast majority
of inconsistencies in a requirements document.
Sec.~\ref{sec:uml} gives an example of to-UML mapping implemented
by the compiler.
Sec.~\ref{sec:ambiguity}
presents a numeric metric that measures ambiguity in a requirements
specification and could be used as a guidance for system analysts.
Sec.~\ref{sec:results} presents empirical results obtained from three
commercial Java web projects over the last few years.
Sec.~\ref{sec:conclusion} presents conclusions and observations made by
the author.

@@

It is an object-oriented CNL that resembles English and, because of that,
can be read and understood by compilers, system analysts, end-users, and
business people. Fig.~\ref{fig:example} presents an example of a short
requirements definition document that describes a calculator that asks its
user for two floating point numbers and returns the result of their
division.

@@

Fig.~\ref{fig:java} contains a close equivalent program in Java.

@@

\begin{ffcode*}{linenos=false}
<class> ::= /[A-Z][A-Za-z]+/
<word> ::= /[a-z]+/
<use-case-ID> ::= /UC[0-9\\.]+/
<flow-ID> ::= /[0-9]+/
<informal> ::= /"([^"]|$\vert$|\\")+"/
\end{ffcode*}

In Fig.~\ref{fig:example}, the entities \ff{SuD}, \ff{Fraction}, \ff{Float}, and
\ff{User} are classes. They start with capital letters and contain only
letters (both lowercase and capital). The \ff{user}, the \ff{fraction}, and
the \ff{quotient} are the variables bound to instances of classes.

Besides, there are a few reserved words, which may have special meaning in
certain places of the text, including \ff{the}, \ff{a}, \ff{includes},
\ff{requires}, etc.
A full list of them is defined below in Sec.~\ref{sec:syntax}.

@@

\begin{ffcode*}{linenos=false}
<SRS> ::= <clause>+
<clause> ::= <class-declaration>
  |$\vert$| <class-construction>
  |$\vert$| <method-declaration>
  |$\vert$| <alternative-flow-declaration>
\end{ffcode*}

@@

\begin{ffcode*}{linenos=false}
<class-declaration> ::= <class>
  <is-a> <parent-class>
<parent-class> ::= <informal> |$\vert$| <class>
<is-a> ::= `is' ( `a' |$\vert$| `an' )
\end{ffcode*}

The second line in Fig.~\ref{fig:example} declares the class \ff{Fraction}
without any super-class but with an informal description
\ff{"math calculator"}.

@@

\begin{ffcode*}{linenos=false}
<class-construction> ::= <class> <includes>
  `:' <slots>
<includes> ::= `includes' |$\vert$| `needs'
<slots> ::= <slot> ( `,' <slot> )*
<slot> ::= <variable> ( `as' <class> )?
<variable> ::= <word> `-s'?
\end{ffcode*}

Fig.~\ref{fig:example} defines constructor arguments only for
the \ff{Fraction} class. The class \ff{User} don't have any arguments. The class
\ff{SuD} is a pre-defined class without arguments. The class \ff{Float} is a
pre-defined class with one argument.

@@

\begin{ffcode*}{linenos=false}
<method-declaration> ::= <use-case-ID> `where'
<class> <signature> `:' <flows>
<signature> ::= ( <word>+ |$\vert$| <informal> )
  <subject>? <using>?
<subject> ::= <class> <binding>?
  |$\vert$| `the' <variable>
<binding> ::= `(' `a' <variable> `)'
<using> ::= `using' <subject>
  ( `and' <subject> )*
<flows> ::= <flow> ( `;' <flow> )+
<flow> ::= <flow-ID> `.'
  (
    `The' <variable> <signature>
    |$\vert$|
    `Fail' `since' <informal>
  )
\end{ffcode*}

@@

\begin{ffcode*}{linenos=false}
<alternative-flow-declaration> ::= <use-case-ID> `/' <flow-ID> `when' <informal> `:' <flows>
\end{ffcode*}

@@

Every method call in every use case should either
\begin{inparaenum}[\itshape a\upshape)]
\item be informal (the 8th, 9th, and 11th lines in the example at Fig.~\ref{fig:example}),
\item formally refer to the method declared in a class, or
\item construct an instance of a class using a built-in method \ff{creates} (the 7th line in the example).
\end{inparaenum}

@@

Every method in the object-oriented way may throw an exception, using a
build-in method \ff{Fail since}. The method has one parameter of the class
\ff{Text}, which explains the reason. In the example in
Fig.~\ref{fig:example}, method ``calculates'' of the class \ff{Fraction} may
throw an exception and it is handled by an alternative flow. In the language, all
exceptions are checked, which means that they should be handled explicitly,
by means of alternative flows of use cases.

@@

When semantic analysis is done the specification looks like an object of
the class \ff{SuD} that encapsulates other objects and methods. This object
hierarchy is convertible to UML diagrams~\citep{uml}. Specification in
Fig.~\ref{fig:example} would be converted to the following UML diagrams
(to save space, the list is limited to the three most interesting
diagrams):

\begin{itemize}
\item Class diagram, in Fig.~\ref{fig:uml-class}
\item Object diagram, in Fig.~\ref{fig:uml-object}
\item Sequence diagram of ``divides two numbers'' method, in Fig.~\ref{fig:uml-seq}
\end{itemize}

@@

Every UML diagram is a formal interoperable document in XMI~\citep{xmi} that
can be rendered in web, in \LaTeX, in PDF, or translated to Java or any
other object-oriented language. Fig.~\ref{fig:xmi} shows an example XMI
for the class diagram.

@@

In Fig.~\ref{fig:example}, ambiguity equals to $3/4$ since there are four
methods in total and three of them are defined informally.

A method is formally defined when its signature consists of one specially
reserved verb \ff{creates} and an optional list of constructor arguments,
as in the seventh line of Fig.~\ref{fig:example}. A formally defined
method doesn't have any ambiguity since it is absolutely clear what is
expected as its input and output. All it does is instantiate a new object
and bind it to a variable. In Java that would mean calling a constructor.

The higher the ambiguity, the more system analysis is required for the
requirements document to make it unambiguously understandable by everybody
in a project. The ultimate goal of a system analyst is to break down
requirements until a target ambiguity is reached. In the commercial projects
explained in Sec.~\ref{sec:results}, the $A$ metric was used for the planning
of RE activities. Every iteration a group of RE specialists had a goal of
decreasing the ambiguity until it reached a given value. It was not clear at
the beginning how much work that would take, but soon the RE team
obtained that knowledge and became capable of estimating
complexity of work using $A$ metric.
